### 2018-01-26 submission/manuscript.Rmd
add ada suggestions methods and discussion

***Obtaining Data Sets:***
The studies used for this meta-analysis were identified through the review articles written by Keku, *et al.* and Vogtmann, *et al.* [@keku_gastrointestinal_2015; @vogtmann_epidemiologic_2016] and additional studies not mentioned in the reviews were obtained based on the authors' knowledge of the literature. Studies that used tissue or feces as their sample source for 454 or Illumina 16S rRNA gene sequencing analysis were included and only data sets that had sequences available for analysis were included. Some studies were excluded because they did not have publicly available sequences or did not have metadata in which the authors were able to share and were excluded. After these filtering steps, the following studies remained: Ahn, *et al.* [@ahn_human_2013], Baxter, *et al.* [@baxter_microbiota-based_2016], Brim, *et al.* [@brim_microbiome_2013], Burns, *et al.* [@burns_virulence_2015], Chen, *et al.*  [@chen_human_2012], Dejea, *et al.* [@dejea_microbiota_2014], Flemer, *et al.* [@flemer_tumour-associated_2017], Geng, *et al.* [@geng_diversified_2013], Hale, *et al.* [@hale_shifts_2017], Kostic, *et al.* [@kostic_genomic_2012], Lu, *et al.* [@lu_mucosal_2016], Sanapareddy, *et al.* [@sanapareddy_increased_2012], Wang, *et al.* [@wang_structural_2012], Weir, *et al.* [@weir_stool_2013], and Zeller, *et al.*  [@zeller_potential_2014]. The Zackular [@zackular_human_2014] study was not included because the 90 individuals analyzed within the study are contained within the larger Baxter study [@baxter_microbiota-based_2016]. After sequence processing all the case samples for the Kostic study had 100 or less sequences remaining and was not used which left a total of 14 studies that analysis could be completed on.

***Data Set Breakdown:***
In total, there were seven studies with only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and Zeller), five studies with only tissue samples (Burns, Dejea, Geng, Lu, Sanapareddy), and two studies with both fecal and tissue samples (Chen and Flemer). The total number of individuals that were analyzed after sequence processing for feces was `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)` [Table 1]. The total number of matched and unmatched tissue samples that were analyzed after sequence processing was `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` [Table 2].

***Sequence Processing:***
For the majority of studies, raw sequences were downloaded from the Sequence Read Archive (SRA) (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained by searching the respective accession number of the study at the following website: http://www.ncbi.nlm.nih.gov/Traces/study/. Of the studies that did not have sequences and metadata on the SRA, data was obtained from DBGap (n = 1, [@ahn_human_2013]) and directly from the authors (n = 4, [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]). Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009] and quality filtering utilized the default methods, used in mothur, for either 454 or Illumina based sequencing. If it was not possible to use the defaults, the stated quality cut-offs, from the study itself, were used instead. Chimeras were identified and removed using VSEARCH [@rognes_vsearch_2016] before *de novo* OTU clustering at 97% similarity was completed using the OptiClust algorithm [@westcott_opticlust_2017].

@@

***Study Analysis Overview:***
$\alpha$-diversity was first assessed for differences between controls, adenoma, and carcinoma using both linear mixed-effect models and RR. $\beta$-diversity was then assessed for each individual study for differences between control-adenoma and control-carcinoma. Next, all common genera were assessed for differences in RR for adenoma and carcinoma and ranked based on P-value. We then built Random Forest models based on all or a selected model, based on the top 5 increased and top 5 decreased RR based on P-value, and these models were trained on one study then tested on the remaining studies, for every study. A similar approach was then applied at the OTU level with the exception that a 10-fold CV over 100 different models, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs. For these OTU-based models, the selected model included all OTUs that had a taxonomic classification to a variable in the top 5 increased and top 5 decreased RR based on P-value. Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power. For comparisons in which normal versus adenoma were made the carcinoma samples were excluded from each respective study. Similarly, for comparisons in which normal versus carcinoma were made the adenoma samples were excluded from each respective study. The data was split between feces and tissue samples. Within the tissue groups the data was further divided between samples from the same individual (matched) and those from different individuals (unmatched). Where applicable for each study, predictions for adenoma and carcinoma were then tested for feces, matched tissue, and unmatched tissue.

***Obtaining Genera Relative Abundance and Selected Models:***
For the genera analysis of the RR, OTUs were added together based on the genus or lowest available taxonomic classification level and the total average counts, for 100 different subsamplings, were collected. The OTU based Random Forest Models using selected OTUS utilized a similar approach except that the OTUs were not aggregated together by taxonomic identity but kept as separate OTUs. OTU Random Forest models using the full community included all OTUs while those for the selected model included only those OTUs that had a taxonomic classification to a variable in the top 5 increased of top 5 decreased RR based on P-value.

***Matched versus Unmatched Tissue Samples:***
In general, tissue samples that had control and lesion samples, that did not belong to the same individual, were classified as unmatched while samples, that belonged to the same individual, were classified as matched. Studies with matched data included Burns, Dejea, Geng, and Lu while those with unmatched data were from Burns, Flemer, Chen, and Sanapareddy. For some studies samples became unmatched due to one of the corresponding matched samples not making it through sequence processing. All samples, from both tissue sample types, were analyzed together for the linear mixed-effect models with samples from the same individual corrected for. For all other analysis, not mentioned herein, matched and unmatched samples were analyzed separately using the statistical approaches mentioned in the Statistical Analysis section.

***Assessing Important Random Forest Model Variables:***
Using Mean Decrease in Accuracy (MDA) the top 10 most important variables were obtained in two different ways depending on whether the model used genera or OTU data. For the genus based models, the number of times that a genus showed up in the top 10 of the training set across each study was counted while, for the OTU based models, the medians for each OTU across 100 different 80/20 splits of the data was generated and the top 10 OTUs then counted for each study. Common taxa, for the OTU based models, were identified by using the lowest classification within the RDP database for each of the specific OTUs obtained from the previous counts and the number of times this classification occurred in the top 10, in each study, was recorded. The two studies that had adenoma tissue were equally divided between matched and unmatched groups and were grouped together for the counting of the top 10 genera and OTUs.

***Reproducible Methods:***
The code and analysis can be found here at  https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017. Unless otherwise mentioned, the accession number for the raw sequences for the studies used in this analysis can be found directly in the respective batch file in the GitHub repository or in the original manuscript.

>>>>

***Obtaining Data Sets:***
The studies used for this meta-analysis were identified through the review articles written by Keku, *et al.* and Vogtmann, *et al.* [@keku_gastrointestinal_2015; @vogtmann_epidemiologic_2016] and additional studies not mentioned in the reviews were obtained based on the authors' knowledge of the literature. Studies that used tissue or feces as their sample source for 454 or Illumina 16S rRNA gene sequencing analysis and had data sets with sequences available for analysis were included. Some studies were excluded because they did not have publicly available sequences or did not have metadata in which the authors were able to share. After these filtering steps, the following studies remained: Ahn, *et al.* [@ahn_human_2013], Baxter, *et al.* [@baxter_microbiota-based_2016], Brim, *et al.* [@brim_microbiome_2013], Burns, *et al.* [@burns_virulence_2015], Chen, *et al.*  [@chen_human_2012], Dejea, *et al.* [@dejea_microbiota_2014], Flemer, *et al.* [@flemer_tumour-associated_2017], Geng, *et al.* [@geng_diversified_2013], Hale, *et al.* [@hale_shifts_2017], Kostic, *et al.* [@kostic_genomic_2012], Lu, *et al.* [@lu_mucosal_2016], Sanapareddy, *et al.* [@sanapareddy_increased_2012], Wang, *et al.* [@wang_structural_2012], Weir, *et al.* [@weir_stool_2013], and Zeller, *et al.*  [@zeller_potential_2014]. The Zackular [@zackular_human_2014] study was not included because the 90 individuals analyzed within the study are contained within the larger Baxter study [@baxter_microbiota-based_2016]. After sequence processing, all the case samples for the Kostic study had 100 or less sequences remaining and was excluded, leaving a total of 14 studies that analysis could be completed on.

***Data Set Breakdown:***
In total, there were seven studies with only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and Zeller), five studies with only tissue samples (Burns, Dejea, Geng, Lu, Sanapareddy), and two studies with both fecal and tissue samples (Chen and Flemer). The total number of individuals analyzed after sequence processing for feces was `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)` [Table 1]. The total number of matched and unmatched tissue samples that were analyzed after sequence processing was `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` [Table 2].

***Sequence Processing:***
For the majority of studies, raw sequences were downloaded from the Sequence Read Archive (SRA) (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata were obtained by searching the respective accession number of the study at the following website: http://www.ncbi.nlm.nih.gov/Traces/study/. Of the studies that did not have sequences and metadata on the SRA, data was obtained from DBGap (n = 1, [@ahn_human_2013]) and directly from the authors (n = 4, [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]). Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009] and quality filtering utilized the default methods for both 454 and Illumina based sequencing. If it was not possible to use the defaults, the stated quality cut-offs, from the study itself, were used instead. Sequences that were made up of an artificial combination of two or more different sequences and commonly known as chimeras were identified and removed using VSEARCH [@rognes_vsearch_2016] before *de novo* OTU clustering at 97% similarity was completed using the OptiClust algorithm [@westcott_opticlust_2017].

@@

***Study Analysis Overview:***
OTU richness, evenness, and Shannon diversity was first assessed for differences between controls, adenoma stage, and carcinoma stage using both linear mixed-effect models and RR. The Bray-Curtis index was used to assess, for each individual study, differences between control-adenoma and control-carcinoma. Next, all common genera were assessed for differences in RR for having an adenoma or carcinoma and ranked based on P-value. We then built Random Forest models based on full or  selected community, based on the top 5 increased and top 5 decreased RR based on P-value, and these models were trained on one study then tested on the remaining studies. This process was repeated for every study in the meta-analysis. A similar approach was then applied at the OTU level with the exception that a 10-fold CV over 100 different models, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs. For these OTU-based models, the selected model included all OTUs that had a taxonomic classification to a taxa in the top 5 increased and top 5 decreased RR based on P-value. Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power. For comparisons in which only control versus adenoma stage were made, the carcinoma samples were excluded from each respective study. Similarly, for comparisons in which control versus carcinoma stage were made the adenoma samples were excluded from each respective study. The data were split between feces and tissue samples. Within the tissue groups the data were further divided between samples from the same individual (matched) and those from different individuals (unmatched). Where applicable for each study, predictions for adenoma and carcinoma stage of disease were then tested for feces, matched tissue, and unmatched tissue.

***Obtaining Genera Relative Abundance and Selected Models:***
For the genera analysis of the RR, OTUs were added together based on the genus or lowest available taxonomic classification level and the total average counts, for 100 different subsamplings, were collected. The OTU based Random Forest Models using selected OTUs utilized a similar approach except that the OTUs were not aggregated together by taxonomic identity but kept as separate OTUs. OTU Random Forest models using the full community included all OTUs while those for the selected model included only those OTUs that had a taxonomic classification to a variable in the top 5 increased of top 5 decreased RR based on P-value.

***Matched versus Unmatched Tissue Samples:***
In general, tissue samples with control and CRC samples from different indivdiuals were classified as unmatched while samples that belonged to the same individual were classified as matched. Studies with matched data included Burns, Dejea, Geng, and Lu while those with unmatched data were from Burns, Flemer, Chen, and Sanapareddy. For some studies samples became unmatched when a corresponding matched sample did not make it through sequence processing. All samples, from both tissue sample types, were analyzed together for the linear mixed-effect models with samples from the same individual corrected for. For all other analysis, not mentioned herein, matched and unmatched samples were analyzed separately using the statistical approaches mentioned in the Statistical Analysis section.

***Assessing Important Random Forest Model Variables:***
Using Mean Decrease in Accuracy (MDA) the top 10 most important variables to the Random Forest model were obtained in two different ways depending on whether the model used genera or OTU data. For the genus based models, the number of times that a genus showed up in the top 10 of the training set across each study was counted while, for the OTU based models, the medians for each OTU across 100 different 80/20 splits of the data was generated and the top 10 OTUs then counted for each study. Common taxa, for the OTU based models, were identified by using the lowest classification within the RDP database for each of the specific OTUs obtained from the previous counts and the number of times this classification occurred in the top 10, in each study, was recorded. The two studies that had adenoma tissue were equally divided between matched and unmatched groups and were grouped together for the counting of the top 10 genera and OTUs.

***Reproducible Methods:***
The code and analysis can be found at https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017. Unless otherwise mentioned, the accession number of raw sequences from the studies used in this analysis can be found directly in the respective batch file in the GitHub repository or in the original manuscript.

### 2018-01-26 submission/manuscript.tex
add ada suggestions methods and discussion

\textbf{\emph{Obtaining Data Sets:}} The studies used for this
meta-analysis were identified through the review articles written by
Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}31,32{]} and
additional studies not mentioned in the reviews were obtained based on
the authors' knowledge of the literature. Studies that used tissue or
feces as their sample source for 454 or Illumina 16S rRNA gene
sequencing analysis were included and only data sets that had sequences
available for analysis were included. Some studies were excluded because
they did not have publicly available sequences or did not have metadata
in which the authors were able to share and were excluded. After these
filtering steps, the following studies remained: Ahn, \emph{et al.}
{[}12{]}, Baxter, \emph{et al.} {[}13{]}, Brim, \emph{et al.} {[}30{]},
Burns, \emph{et al.} {[}16{]}, Chen, \emph{et al.} {[}14{]}, Dejea,
\emph{et al.} {[}24{]}, Flemer, \emph{et al.} {[}18{]}, Geng, \emph{et
al.} {[}23{]}, Hale, \emph{et al.} {[}27{]}, Kostic, \emph{et al.}
{[}33{]}, Lu, \emph{et al.} {[}26{]}, Sanapareddy, \emph{et al.}
{[}25{]}, Wang, \emph{et al.} {[}15{]}, Weir, \emph{et al.} {[}21{]},
and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}34{]} study was not
included because the 90 individuals analyzed within the study are
contained within the larger Baxter study {[}13{]}. After sequence
processing all the case samples for the Kostic study had 100 or less
sequences remaining and was not used which left a total of 14 studies
that analysis could be completed on.

\textbf{\emph{Data Set Breakdown:}} In total, there were seven studies
with only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and
Zeller), five studies with only tissue samples (Burns, Dejea, Geng, Lu,
Sanapareddy), and two studies with both fecal and tissue samples (Chen
and Flemer). The total number of individuals that were analyzed after
sequence processing for feces was 1737 {[}Table 1{]}. The total number
of matched and unmatched tissue samples that were analyzed after
sequence processing was 492 {[}Table 2{]}.

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained by searching the respective accession number
of the study at the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
(n = 1, {[}12{]}) and directly from the authors (n = 4,
{[}18,21,25,27{]}). Each study was processed using the mothur (v1.39.3)
software program {[}35{]} and quality filtering utilized the default
methods, used in mothur, for either 454 or Illumina based sequencing. If
it was not possible to use the defaults, the stated quality cut-offs,
from the study itself, were used instead. Chimeras were identified and
removed using VSEARCH {[}36{]} before \emph{de novo} OTU clustering at
97\% similarity was completed using the OptiClust algorithm {[}37{]}.

@@

\textbf{\emph{Study Analysis Overview:}} \(\alpha\)-diversity was first
assessed for differences between controls, adenoma, and carcinoma using
both linear mixed-effect models and RR. \(\beta\)-diversity was then
assessed for each individual study for differences between
control-adenoma and control-carcinoma. Next, all common genera were
assessed for differences in RR for adenoma and carcinoma and ranked
based on P-value. We then built Random Forest models based on all or a
selected model, based on the top 5 increased and top 5 decreased RR
based on P-value, and these models were trained on one study then tested
on the remaining studies, for every study. A similar approach was then
applied at the OTU level with the exception that a 10-fold CV over 100
different models, based on random 80/20 splitting of the data, was used
to generate a range of expected AUCs. For these OTU-based models, the
selected model included all OTUs that had a taxonomic classification to
a variable in the top 5 increased and top 5 decreased RR based on
P-value. Finally, the power of each study was assessed for an effect
size ranging from 1\% to 30\% and an estimated sample size, for these
effect sizes, was generated based on 80\% power. For comparisons in
which normal versus adenoma were made the carcinoma samples were
excluded from each respective study. Similarly, for comparisons in which
normal versus carcinoma were made the adenoma samples were excluded from
each respective study. The data was split between feces and tissue
samples. Within the tissue groups the data was further divided between
samples from the same individual (matched) and those from different
individuals (unmatched). Where applicable for each study, predictions
for adenoma and carcinoma were then tested for feces, matched tissue,
and unmatched tissue.

\textbf{\emph{Obtaining Genera Relative Abundance and Selected Models:}}
For the genera analysis of the RR, OTUs were added together based on the
genus or lowest available taxonomic classification level and the total
average counts, for 100 different subsamplings, were collected. The OTU
based Random Forest Models using selected OTUS utilized a similar
approach except that the OTUs were not aggregated together by taxonomic
identity but kept as separate OTUs. OTU Random Forest models using the
full community included all OTUs while those for the selected model
included only those OTUs that had a taxonomic classification to a
variable in the top 5 increased of top 5 decreased RR based on P-value.

\textbf{\emph{Matched versus Unmatched Tissue Samples:}} In general,
tissue samples that had control and lesion samples, that did not belong
to the same individual, were classified as unmatched while samples, that
belonged to the same individual, were classified as matched. Studies
with matched data included Burns, Dejea, Geng, and Lu while those with
unmatched data were from Burns, Flemer, Chen, and Sanapareddy. For some
studies samples became unmatched due to one of the corresponding matched
samples not making it through sequence processing. All samples, from
both tissue sample types, were analyzed together for the linear
mixed-effect models with samples from the same individual corrected for.
For all other analysis, not mentioned herein, matched and unmatched
samples were analyzed separately using the statistical approaches
mentioned in the Statistical Analysis section.

\textbf{\emph{Assessing Important Random Forest Model Variables:}} Using
Mean Decrease in Accuracy (MDA) the top 10 most important variables were
obtained in two different ways depending on whether the model used
genera or OTU data. For the genus based models, the number of times that
a genus showed up in the top 10 of the training set across each study
was counted while, for the OTU based models, the medians for each OTU
across 100 different 80/20 splits of the data was generated and the top
10 OTUs then counted for each study. Common taxa, for the OTU based
models, were identified by using the lowest classification within the
RDP database for each of the specific OTUs obtained from the previous
counts and the number of times this classification occurred in the top
10, in each study, was recorded. The two studies that had adenoma tissue
were equally divided between matched and unmatched groups and were
grouped together for the counting of the top 10 genera and OTUs.

\textbf{\emph{Reproducible Methods:}} The code and analysis can be found
here at
\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.
Unless otherwise mentioned, the accession number for the raw sequences
for the studies used in this analysis can be found directly in the
respective batch file in the GitHub repository or in the original
manuscript.

>>>>

\textbf{\emph{Obtaining Data Sets:}} The studies used for this
meta-analysis were identified through the review articles written by
Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}31,32{]} and
additional studies not mentioned in the reviews were obtained based on
the authors' knowledge of the literature. Studies that used tissue or
feces as their sample source for 454 or Illumina 16S rRNA gene
sequencing analysis and had data sets with sequences available for
analysis were included. Some studies were excluded because they did not
have publicly available sequences or did not have metadata in which the
authors were able to share. After these filtering steps, the following
studies remained: Ahn, \emph{et al.} {[}12{]}, Baxter, \emph{et al.}
{[}13{]}, Brim, \emph{et al.} {[}30{]}, Burns, \emph{et al.} {[}16{]},
Chen, \emph{et al.} {[}14{]}, Dejea, \emph{et al.} {[}24{]}, Flemer,
\emph{et al.} {[}18{]}, Geng, \emph{et al.} {[}23{]}, Hale, \emph{et
al.} {[}27{]}, Kostic, \emph{et al.} {[}33{]}, Lu, \emph{et al.}
{[}26{]}, Sanapareddy, \emph{et al.} {[}25{]}, Wang, \emph{et al.}
{[}15{]}, Weir, \emph{et al.} {[}21{]}, and Zeller, \emph{et al.}
{[}17{]}. The Zackular {[}34{]} study was not included because the 90
individuals analyzed within the study are contained within the larger
Baxter study {[}13{]}. After sequence processing, all the case samples
for the Kostic study had 100 or less sequences remaining and was
excluded, leaving a total of 14 studies that analysis could be completed
on.

\textbf{\emph{Data Set Breakdown:}} In total, there were seven studies
with only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and
Zeller), five studies with only tissue samples (Burns, Dejea, Geng, Lu,
Sanapareddy), and two studies with both fecal and tissue samples (Chen
and Flemer). The total number of individuals analyzed after sequence
processing for feces was 1737 {[}Table 1{]}. The total number of matched
and unmatched tissue samples that were analyzed after sequence
processing was 492 {[}Table 2{]}.

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata were obtained by searching the respective accession number
of the study at the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
(n = 1, {[}12{]}) and directly from the authors (n = 4,
{[}18,21,25,27{]}). Each study was processed using the mothur (v1.39.3)
software program {[}35{]} and quality filtering utilized the default
methods for both 454 and Illumina based sequencing. If it was not
possible to use the defaults, the stated quality cut-offs, from the
study itself, were used instead. Sequences that were made up of an
artificial combination of two or more different sequences and commonly
known as chimeras were identified and removed using VSEARCH {[}36{]}
before \emph{de novo} OTU clustering at 97\% similarity was completed
using the OptiClust algorithm {[}37{]}.

@@

\textbf{\emph{Study Analysis Overview:}} OTU richness, evenness, and
Shannon diversity was first assessed for differences between controls,
adenoma stage, and carcinoma stage using both linear mixed-effect models
and RR. The Bray-Curtis index was used to assess, for each individual
study, differences between control-adenoma and control-carcinoma. Next,
all common genera were assessed for differences in RR for having an
adenoma or carcinoma and ranked based on P-value. We then built Random
Forest models based on full or selected community, based on the top 5
increased and top 5 decreased RR based on P-value, and these models were
trained on one study then tested on the remaining studies. This process
was repeated for every study in the meta-analysis. A similar approach
was then applied at the OTU level with the exception that a 10-fold CV
over 100 different models, based on random 80/20 splitting of the data,
was used to generate a range of expected AUCs. For these OTU-based
models, the selected model included all OTUs that had a taxonomic
classification to a taxa in the top 5 increased and top 5 decreased RR
based on P-value. Finally, the power of each study was assessed for an
effect size ranging from 1\% to 30\% and an estimated sample size, for
these effect sizes, was generated based on 80\% power. For comparisons
in which only control versus adenoma stage were made, the carcinoma
samples were excluded from each respective study. Similarly, for
comparisons in which control versus carcinoma stage were made the
adenoma samples were excluded from each respective study. The data were
split between feces and tissue samples. Within the tissue groups the
data were further divided between samples from the same individual
(matched) and those from different individuals (unmatched). Where
applicable for each study, predictions for adenoma and carcinoma stage
of disease were then tested for feces, matched tissue, and unmatched
tissue.

\textbf{\emph{Obtaining Genera Relative Abundance and Selected Models:}}
For the genera analysis of the RR, OTUs were added together based on the
genus or lowest available taxonomic classification level and the total
average counts, for 100 different subsamplings, were collected. The OTU
based Random Forest Models using selected OTUs utilized a similar
approach except that the OTUs were not aggregated together by taxonomic
identity but kept as separate OTUs. OTU Random Forest models using the
full community included all OTUs while those for the selected model
included only those OTUs that had a taxonomic classification to a
variable in the top 5 increased of top 5 decreased RR based on P-value.

\textbf{\emph{Matched versus Unmatched Tissue Samples:}} In general,
tissue samples with control and CRC samples from different indivdiuals
were classified as unmatched while samples that belonged to the same
individual were classified as matched. Studies with matched data
included Burns, Dejea, Geng, and Lu while those with unmatched data were
from Burns, Flemer, Chen, and Sanapareddy. For some studies samples
became unmatched when a corresponding matched sample did not make it
through sequence processing. All samples, from both tissue sample types,
were analyzed together for the linear mixed-effect models with samples
from the same individual corrected for. For all other analysis, not
mentioned herein, matched and unmatched samples were analyzed separately
using the statistical approaches mentioned in the Statistical Analysis
section.

\textbf{\emph{Assessing Important Random Forest Model Variables:}} Using
Mean Decrease in Accuracy (MDA) the top 10 most important variables to
the Random Forest model were obtained in two different ways depending on
whether the model used genera or OTU data. For the genus based models,
the number of times that a genus showed up in the top 10 of the training
set across each study was counted while, for the OTU based models, the
medians for each OTU across 100 different 80/20 splits of the data was
generated and the top 10 OTUs then counted for each study. Common taxa,
for the OTU based models, were identified by using the lowest
classification within the RDP database for each of the specific OTUs
obtained from the previous counts and the number of times this
classification occurred in the top 10, in each study, was recorded. The
two studies that had adenoma tissue were equally divided between matched
and unmatched groups and were grouped together for the counting of the
top 10 genera and OTUs.

\textbf{\emph{Reproducible Methods:}} The code and analysis can be found
at
\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.
Unless otherwise mentioned, the accession number of raw sequences from
the studies used in this analysis can be found directly in the
respective batch file in the GitHub repository or in the original
manuscript.
